Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-10-29 manu-draft.qmd
started results text

load(file = here('data/gamfit.RData'))
load(file = here('tabs/temptrndtab.RData'))
load(file = here('tabs/saltrndtab.RData'))
load(file = here('tabs/mixdaytab.RData'))
```

@@

A fundamental challenge in describing niche space is identifying the boundaries for optimal conditions. In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985]. These species co-occur often in mixed beds, although differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligohaline portions of Tampa Bay. Conversely, turtle grass less tolerant of low salinity and is most often found in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].

@@

For @fig-meteowqraw, salinity shows much higher inter-annual variability, but lower intra-year variation among stations. Trends are similar for bottom vs top, temperature shows stronger trend than salinity.

Kendall test results are shown in @fig-kendall.

Linear trends by dataset

>>>>

source(here('R/funcs.R'))

load(file = here('data/speidat.RData'))
load(file = here('data/gamfit.RData'))
load(file = here('tabs/temptrndtab.RData'))
load(file = here('tabs/saltrndtab.RData'))
load(file = here('tabs/mixdaytab.RData'))
```

@@

A fundamental challenge in describing niche space is identifying the boundaries for optimal conditions. In Tampa Bay, the three dominant seagrass species are *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985]. These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay. Conversely, turtle grass is less tolerant of low salinity and is most often found in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].

@@

```{r}
data(speidat)
airtemp <- speidat %>% 
  select(yr, tavg_c) %>% 
  summarise(temp = mean(tavg_c), .by = yr)
airtempmod <- lm(temp ~ yr, airtemp)
airtempslo <- round(coefficients(summary(airtempmod))[2, 1], 2)
airtemppvl <- p_txt(coefficients(summary(airtempmod))[2, 4])
airtemprsq <- round(summary(airtempmod)$adj.r.squared, 2)
airtempprd <- predict(airtempmod, se = T)
airtempstr <- paste0(round(airtempprd$fit[1], 1), ' (+/-', round(airtempprd$se.fit[1], 2), ' st. err.)')
airtempend <- paste0(round(airtempprd$fit[nrow(airtemp)], 1), ' (+/-', round(airtempprd$se.fit[nrow(airtemp)], 2), ' st. err.)')
airtempdif <- round(airtempprd$fit[nrow(airtemp)]- airtempprd$fit[1], 1)

prp <- speidat %>% 
  select(yr, precip_mm) %>% 
  summarise(prptot = sum(precip_mm), .by = yr)
prpmod <- lm(prptot ~ yr, prp)
prpslo <- round(coefficients(summary(prpmod))[2, 1], 2)
prppvl <- p_txt(coefficients(summary(prpmod))[2, 4])
prprsq <- round(summary(prpmod)$adj.r.squared, 2)
prpprd <- predict(prpmod, se = T)
prpstr <- paste0(round(prpprd$fit[1], 1), ' (+/-', round(prpprd$se.fit[1], 2), ' st. err.)')
prpend <- paste0(round(prpprd$fit[nrow(prp)], 1), ' (+/-', round(prpprd$se.fit[nrow(prp)], 2), ' st. err.)')
prpdif <- round(prpprd$fit[nrow(prp)]- prpprd$fit[1], 1)

# load(file = url("https://github.com/tbep-tech/load-estimates/raw/main/data/totanndat.RData"))
# 
# # hydro load, otb, mtb, ltb, hb
# hydrodat <- totanndat %>% 
#   select(year, bay_segment, hy_load) %>% 
#   filter(bay_segment %in% c('Hillsborough Bay', 'Old Tampa Bay', 'Lower Tampa Bay', 'Middle Tampa Bay')) %>% 
#   summarise(
#     hy_load = sum(hy_load) / 1e3, 
#     .by = 'year'
#   ) 
```

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature an precipitation [@fig-meteowqraw]. Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$. Total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.  The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).  The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.  The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay ([@fig-meteowqraw]).  Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).  Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.  As expected, intra-annual variability of water temperature was much higher than salinity following the seasonal variation in temperature ([@fig-meteowqraw]).  This variation may also be related to differences among the monitoring stations in each segment.  Statistically significant increasing temperature trends were also observed for the FIM datasets from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).  No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).  However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.  Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).

The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall).  All stations in both the top and bottom of the water column had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a). Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figures [-@fig-kendall]b,c), although some variation was observed between bay segments.  Temperature trends were more often observed to occur in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more significant trends during the spring.  Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more significantly decreasing trends than the other bay segments. Again, differences between top or bottom trends did not differ seasonally.
